Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-02-16 paper/zhang-cook-laa-langrene-menendez.Rmd
fix kol-geo-pca plot and writing
write up torus section and update  vimeo links

```{r toy-pca-animated, fig.cap = "Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \\url{https://vimeo.com/user132007430/review/504242845/b73f37175a}."}
```

@@

```{r toy-tour, fig.cap="Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \\ref{fig:toy-pca} and \\ref{fig:toy-pca-animated}. The full video of the animation can be found at \\url{https://vimeo.com/user132007430/review/504328122/9be84db563}."}
```

@@

\textcolor{red}{Something to fill here ...}

@@

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. \textcolor{red}{[A sentence for PD]}. Comparing the amount of random search done by SA and SAJO, it is surprising that SAJO does not carry as many samples as SA. Combining the insights from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SAJO first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SAJO has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

>>>>

```{r toy-pca-animated, fig.cap = "Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \\url{https://vimeo.com/504242845}."}
```

@@

```{r toy-tour, fig.cap="Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \\ref{fig:toy-pca} and \\ref{fig:toy-pca-animated}. The full video of the animation can be found at \\url{https://vimeo.com/512885379}."}
```

@@

While the previous few examples have looked at the space of 1D bases in a unit sphere, this section visualises the space of 2D bases. Recall that the columns in a 2D basis are orthogonal to each other, the space of $p \times 2$ bases is a $p$-D torus. If take $p = 3$, one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality of the columns in the bases. Two paths optimised by SA and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous few 1D space plots. The final basis found by PD and SA are shown in a larger shape  and printed below respectively:

```{r}
holes_2d_geo_3var %>% get_best() %>% pull(basis) %>% .[[1]]
```

```{r}
holes_2d_better_3var %>% get_best() %>% pull(basis) %>% .[[1]]
```

Both optimisers have found the third variable in the first column, but the second variable in the second column differs by a sign.One would expect to see this in the torus plot as the final bases match each other when projected onto one torus circle (due to the same sign in the first column) and be symmetric when projected onto the other (due the the sign difference in the second column). In Figure \ref{fig:toy-torus}, this can be seen most clearly in frame 5 where the two circles are rotated into a line from our view.

```{r toy-torus, fig.cap="Six frames selected from the torus view of 2D basis space. The grey points are randomly generated 2D basis and the two paths are optimised by PD (brown) and SA (green) for a three-variable 2D projection problem using the holes index. The grey points show the space of the 2D basis as a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}."}
```

@@

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. Pseudo Derivative (PD) quickly gets stuck near the starting position. Comparing the amount of random search done by SA and SAJO, it is surprising that SAJO does not carry as many samples as SA. Combining the insights from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SAJO first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SAJO has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

### 2021-02-16 paper/zhang-cook-laa-langrene-menendez.tex
fix kol-geo-pca plot and writing
write up torus section and update  vimeo links

\caption{Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}.}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

@@

\caption{Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \ref{fig:toy-pca} and \ref{fig:toy-pca-animated}. The full video of the animation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}.}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

@@

\textcolor{red}{Something to fill here ...}

>>>>

\caption{Six frames selected from the animation of search paths in the PCA-projected basis space. With animation, the start and end position of each optimisation routine is clearly identified. The full video of the animation can be found at \url{https://vimeo.com/504242845}.}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

@@

\caption{Six frames selected from the tour view in the original 5D space for the same 1D projection problem as Figure \ref{fig:toy-pca} and \ref{fig:toy-pca-animated}. The full video of the animation can be found at \url{https://vimeo.com/512885379}.}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

@@

While the previous few examples have looked at the space of 1D bases in
a unit sphere, this section visualises the space of 2D bases. Recall
that the columns in a 2D basis are orthogonal to each other, the space
of \(p \times 2\) bases is a \(p\)-D torus. If take \(p = 3\), one would
see a classical 3D torus shape as shown by the grey points in Figure
\ref{fig:toy-torus}. The two circles of the torus can be observed to be
perpendicular to each other and this can be linked back to the
orthogonality of the columns in the bases. Two paths optimised by SA and
PD are plotted on top of the torus and coloured in green and brown
respectively to match the previous few 1D space plots. The final basis
found by PD and SA are shown in a larger shape and printed below
respectively:

\begin{Schunk}
\begin{Soutput}
#>              [,1]        [,2]
#> [1,]  0.001196285  0.03273881
#> [2,] -0.242432715  0.96965761
#> [3,] -0.970167484 -0.24226493
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{Soutput}
#>             [,1]         [,2]
#> [1,]  0.05707994 -0.007220138
#> [2,] -0.40196202 -0.915510160
#> [3,] -0.91387549  0.402230054
\end{Soutput}
\end{Schunk}

Both optimisers have found the third variable in the first column, but
the second variable in the second column differs by a sign.One would
expect to see this in the torus plot as the final bases match each other
when projected onto one torus circle (due to the same sign in the first
column) and be symmetric when projected onto the other (due the the sign
difference in the second column). In Figure \ref{fig:toy-torus}, this
can be seen most clearly in frame 5 where the two circles are rotated
into a line from our view.

{\centering \includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/toy-torus-1}

\caption{Six frames selected from the torus view of 2D basis space. The grey points are randomly generated 2D basis and the two paths are optimised by PD (brown) and SA (green) for a three-variable 2D projection problem using the holes index. The grey points show the space of the 2D basis as a torus. The full video of the animation can be found at \url{https://vimeo.com/512882784}.}\label{fig:toy-torus}
\end{figure}
\end{Schunk}
